Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-06-13 demystify.tex
updates

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for international students, but I have expanded it to include information that might also be useful for \emph{domestic students}.
Moreover, while this handbook is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US (and universities in other countries).

@@

\textbf{Advice}: If you haven't taken the GRE or have bad scores, then don't waste time (re)taking it.  However, if you took it and have really good scores then you can include and talk about them in your application (but don't expect them to make much difference). But if your scores are bad, then you should not include them in your application, which can be a \red{red flag}.

\paragraph{English Tests} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English tests.

@@

\section{CV/Resume}
The CV/Resume provides a summary of the applicant's achievements.

@@

\paragraph{Not getting interviews} While it is generally good to get an interview, not getting one \emph{does not} mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most of our admitted students with TA (\autoref{sec:ta}) do not go through interviews.

@@

Not every faculty can serve as your official PhD adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, and research professors. Here is a primer on these terminologies.

\paragraph{Types of Faculty} There are typically three types of faculty: tenure-line, teaching, and research.  The main difference is their focus: research, teaching, or both.  The type of faculty can affect their ability to serve as your adviser.

@@

Note that non-CS profs (even in tenure-line ones) often cannot solely advise and graduate PhD students (see \autoref{sec:related-fields}).

@@

A common question is whether you should mention that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group (e.g., in your SOP).
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{indeed be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academic.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

It is understandable that you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$30000 and a 3-month summer of \$10000 (a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

In the end, the total budget comes out to be \$72K/year to support a PhD student. \emph{The summary is that over your 5--6 years of your PhD, you cost about \$400K, and while your stipend is X, your adviser probably pays 2X for you}. But of course, the nicest thing is that you do not have to pay for any of this! You get to gain the knowledge, do research, travel, (and don't even have to buy any research or computer equipment as mentioned in \autoref{sec:buying-equipment}) and also get paid!

@@

For start date or TA assignment (e.g., TA'ing a specific course), you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely non-negotiable.  However, you can ask for things such as computer equipment (\autoref{sec:buying-equipment}), books, (and maybe even furniture), and a conference travel budget.

@@

I've noticed many students, especially international, face a dilemma when they commit to a graduate offer but then receive a better one, often discussed in online forums. Advice often suggests it's okay to switch, using reasons like not having a strong relationship with the prof. or prioritizing personal benefit.

@@

@@

I was also a member of the large VietPhD group on Facebook and browsed Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most active participants are students in non-CS fields or not in the US, and their answers are unfortunately not always accurate and sometimes lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider prospective.

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The document was initially for international students but has been generalized to all students.
I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.

\paragraph{About Me} I am an assistant professor in the CS dept at George Mason University (GMU). Before GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at GMU (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

>>>>

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for international students, but I have expanded it to include information that might also be useful for \emph{domestic students}.
Moreover, while this handbook is primarily intended for students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US (and universities in other countries).

@@

\textbf{Advice}: If you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. It does not help much!

However, if you took it and have really good scores then it might be worth it to include and talk about them in your application (but don't expect them to make much difference). But if your scores are bad, then you should not include them in your application, which can be a \red{red flag}.

\paragraph{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English tests.

@@

\section{CV/Resume}
The CV/Resume provides a summary of the applicant's achievements. Most schools require you to upload your CV with your application.

@@

\paragraph{Not getting interviews} While it is generally good to get an interview, not getting one \emph{does not} mean you're out.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most admitted students with TA (\autoref{sec:ta}) do not go through interviews.

@@

\emph{Not} every faculty can serve as your official PhD adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, and research professors. Here is a primer on these terminologies.

\paragraph{Types of Faculty} There are typically two types of faculty: tenure-line and teaching.  The main difference is their focus: research, teaching, or both.  The type of faculty can affect their ability to serve as your adviser.

@@

Note that at many universities do not allow \emph{non-CS profs}, even in tenure-line ones, to solely advise and graduate PhD students in CS (see \autoref{sec:related-fields}).

@@

A common question is whether you should mention, e.g., in your SOP, that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group.
This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences.
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{indeed be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and \href{https://www.gmu.edu/news/2022-09/mason-now-top-10-public-university-diversity-innovation-and-cybersecurity-education-us}{GMU} often promotes itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

Understandably, you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$30K and a 3-month summer of \$10K (a third of the 9-month stipend). I will also use GMU tuition rate of about \$15K/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

In the end, the total budget comes out to be \$73K/year to support a PhD student. \emph{The summary is that over your 5--6 years of your PhD, you cost about \$400K, and while your stipend is X, your adviser probably pays 2X for you}. But of course, the nicest thing is that you do not have to pay for any of this! You get to gain the knowledge, do research, travel, (and don't even have to buy any research or computer equipment as mentioned in \autoref{sec:buying-equipment}) and also get paid!

@@

For a specific start date or TA assignment (e.g., TA'ing a particular course), you can ask for it. Also, there is typically no moving allowance for PhD students. In short, standard things set by the university or department are unlikely negotiable.  However, you can ask for things such as computer equipment (\autoref{sec:buying-equipment}), books, (and maybe even furniture), and a conference travel budget.

@@

I've seen many students, especially international, face a dilemma when they \emph{commit} to a graduate offer but then receive a better one. Advice given in online forums is often along the line that it's okay to switch, using reasons like you haven't yet had a strong relationship with the prof. or you should prioritize personal benefit.

@@

\begin{commentbox}[Vu:]
    My lab has various activities together (see \href{https://photos.app.goo.gl/LFtbqQUuznq9eiL7A}{here}). For example, we always have a Thanksgiving get-together at my house. Many faculty I know have similar activities with their lab members. 
  \end{commentbox}

@@

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most active participants are students in non-CS fields or not in the US, and their answers are unfortunately not always accurate and sometimes lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider perspective.

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The document was initially intended for international students but has been generalized to all students.
I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (GMU). Before GMU, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at GMU and UNL for many years.  Currently I serve as the program director of the MS program in Software Engineering at GMU (thus also have experience with the MS admission process---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.
